HO unification from object language to meta language

### 2024-02-19 paper/main.tex
add paper conf template

\documentclass{rapport}

\usepackage{myTools}

\usepackage{macros}

\author{Enrico Tassi, Davide Fissore}
\title{HO unification from object language to meta language} 
\date{Month XX, 202Y}

\xspaceaddexceptions{]\}}

@@

\newcommand{\library}[1]{\textit{#1}\xspace}
\def\stdpp{\library{stdpp}}
\def\iris{\library{iris}}
% \tableofcontents

@@

\begin{comment}
\begin{abstract}

  % Unification is a fundamental process behind goal resolution in logic
  % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
  % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
  % substitution is a mapping from unification variables to terms and it is the most
  % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
  % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
  % Unification has been widely studied in the 90s [here the refs]. In this paper we
  % want to give a framework to reason about unification in meta-programming.
  Meta-programming \cite{hill1994} is a programming technique in which a
  program, referred to as meta language (\ml), can treat an other program, knwon
  as the object language (\ol) as its data. This programming approach is used in
  various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
  capable of automating proof resolution. In many of the
  previously mentioned works, the equational theory expressed by the \ol is
  weaker than the \ml one. Consequently, no effort is required to adjust the
  unification algorithm of the \ml to represent the unification process of 
  the \ol: the former algorithm encompassing the second one. In our
  specific scenario, however, we work with an \ol with a richer equational
  theory than our \ml. This paper introduces a framework that enables the \ml to
  comprehend the unification of the \ol in the context of higher-order 
  unification. Practical applications of these findings will be provided.

\end{abstract}
\end{comment}

@@

@@

% \begin{elpicode}
%   kind term type.
%   type const gref -> term.
%   type app list term -> term.
%   type fun (term -> term) -> term.
%   type prod (term -> term) -> term.
% \end{elpicode}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Examples
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{mintedTcolorbox}{elpi}
  kind tm type.
  type app list tm -> tm.
  type lam (tm -> tm) -> tm.
  type c string -> tm.
  type uv  nat -> tm.
\end{mintedTcolorbox}

@@

The result of this translation of terms inside the \ml causes a certain lack of
powerfulness while symbolizing higher-order variables. Recall that we are
considering a \ml capable to deal with higher-order variables, however, the
sub-term \elpiIn{app[uv 1, x]} is not expressed into the canonical form where a
higher-order variable of the \ml is in the pattern fragment \cite{miller1991},
i.e. a variable applied to distinct names. Therefore, we need to preprocess the
received unification problem $t_1 = t_2$ by 
\begin{inlinelist}
  \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
  \ml
  \item finding a valid substitution for $t_1'$ and $t_2'$ 
  \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
  is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
\end{inlinelist}.

@@

\begin{mintedTcolorbox}{elpi}
  kind ml.tm type.
  type ml.app list ml.tm -> ml.tm.
  type ml.lam (ml.tm -> ml.tm) -> ml.tm.
  type ml.c string -> ml.tm.
  type ml.uv nat -> list ml.tm -> ml.tm.
\end{mintedTcolorbox}

@@

\printbibliography

@@

\begin{mintedTcolorbox}[escapeinside=&&]{coq}
  Inductive sig (A : Type) (P : A -> Prop) : Type := ...

@@

  Instance decide_rel: forall (A B : Type) (R : A -> B -> Prop),
    RelDecision R -> forall (x : A) (y : B), Decision (R x y). Admitted.
  Instance True_pi : ProofIrrel True. Admitted.
  Instance sig_eq_dec: forall (A : Type) (P : A -> Prop),
    (forall x, ProofIrrel (P x)) -> RelDecision (@eq A) ->
    RelDecision (@eq (sig A P)). Admitted.
\end{mintedTcolorbox}

@@

\begin{mintedTcolorbox}{elpi}
  type tc-Decision term -> term -> o.
  type tc-RelDecision term -> term -> term -> term -> o.
  type tc-ProofIrrel term -> term -> o.

  tc-ProofIrrel (`True`) (`True_pi`).
  tc-Decision (app [R, X, Y]) 
    (app [`decide_rel`, A, B, R, P, X, Y]) :-
    tc-RelDecision A B R P.
  tc-RelDecision (app [`sig`, A, P]) 
    (app [`sig`, A, P]) 
    (app [`eq`, app [`sig`, A, P]]) 
    (app [`sig_eq_dec`, A, P, P1, P2]) :-
    pi-decl c0 `x` A =>
      tc-ProofIrrel (app [P, c0]) (app [P1, c0]), 
    tc-RelDecision A A (app [`eq`, A]) P2.
\end{mintedTcolorbox}

@@

>>>>

\documentclass[acmengage]{acmart}

\usepackage{myTools}
% \addbibresource{bib.bib}
\usepackage{macros}

% TODO: set this fields
\setcopyright{cc}
\setcctype{by}
\copyrightyear{2024}
\acmYear{XXXX 2024}
\acmBooktitle{YYY}
\acmDOI{ZZZZZZZZZZZZ}

% \xspaceaddexceptions{]\}}

@@

\newcommand{\library}[1]{\textit{#1}\xspace}
\def\stdpp{\library{stdpp}}
\def\iris{\library{iris}}

@@

\title{HO unification from object language to meta language}

\author{Enrico Tassi}
\email{enrico.tassi@inria.fr}
\affiliation{%
  \institution{Université Côte d'Azur, Inria}
  % \city{Nice}
  \country{France}}

\author{Davide Fissore}
\email{davide.fissore@inria.fr}
\affiliation{%
  \institution{Université Côte d'Azur, Inria}
  % \city{Nice}
  \country{France}}

@@

\keywords{Logic Programming, Meta-Programming, Higher-Order Unification, Proof Automation}

% \begin{comment}
% \begin{abstract}

%   % Unification is a fundamental process behind goal resolution in logic
%   % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
%   % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
%   % substitution is a mapping from unification variables to terms and it is the most
%   % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
%   % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
%   % Unification has been widely studied in the 90s [here the refs]. In this paper we
%   % want to give a framework to reason about unification in meta-programming.
%   Meta-programming \cite{hill1994} is a programming technique in which a
%   program, referred to as meta language (\ml), can treat an other program, knwon
%   as the object language (\ol) as its data. This programming approach is used in
%   various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
%   capable of automating proof resolution. In many of the
%   previously mentioned works, the equational theory expressed by the \ol is
%   weaker than the \ml one. Consequently, no effort is required to adjust the
%   unification algorithm of the \ml to represent the unification process of 
%   the \ol: the former algorithm encompassing the second one. In our
%   specific scenario, however, we work with an \ol with a richer equational
%   theory than our \ml. This paper introduces a framework that enables the \ml to
%   comprehend the unification of the \ol in the context of higher-order 
%   unification. Practical applications of these findings will be provided.

% \end{abstract}
% \end{comment}

@@

% % \begin{elpicode}
% %   kind term type.
% %   type const gref -> term.
% %   type app list term -> term.
% %   type fun (term -> term) -> term.
% %   type prod (term -> term) -> term.
% % \end{elpicode}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % Examples
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{minted}{elpi}
  kind tm type.
  type app list tm -> tm.
  type lam (tm -> tm) -> tm.
  type c string -> tm.
  type uv  nat -> tm.
\end{minted}

@@

The result of this translation of terms inside the \ml causes a certain lack of
powerfulness while symbolizing higher-order variables. Recall that we are
considering a \ml capable to deal with higher-order variables, however, the
sub-term \elpiIn{app[uv 1, x]} is not expressed into the canonical form where a
higher-order variable of the \ml is in the pattern fragment \cite{miller1991},
i.e. a variable applied to distinct names. Therefore, we need to preprocess the
received unification problem $t_1 = t_2$ by 
% \begin{inlinelist}
%   \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
%   \ml
%   \item finding a valid substitution for $t_1'$ and $t_2'$ 
%   \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
%   is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
% \end{inlinelist}.

@@

\begin{minted}{elpi}
  kind ml.tm type.
  type ml.app list ml.tm -> ml.tm.
  type ml.lam (ml.tm -> ml.tm) -> ml.tm.
  type ml.c string -> ml.tm.
  type ml.uv nat -> list ml.tm -> ml.tm.
\end{minted}

@@

% \printbibliography

@@

\begin{minted}[escapeinside=&&]{coq}
  Inductive sig (A : Type) (P : A -> Prop) : Type := ...

@@

  Instance decide_rel: forall (A B : Type) (R : A -> B -> Prop),
    RelDecision R -> forall (x : A) (y : B), Decision (R x y). Admitted.
  Instance True_pi : ProofIrrel True. Admitted.
  Instance sig_eq_dec: forall (A : Type) (P : A -> Prop),
    (forall x, ProofIrrel (P x)) -> RelDecision (@eq A) ->
    RelDecision (@eq (sig A P)). Admitted.
\end{minted}

@@

\begin{minted}{elpi}
  type tc-Decision term -> term -> o.
  type tc-RelDecision term -> term -> term -> term -> o.
  type tc-ProofIrrel term -> term -> o.

  tc-ProofIrrel (`True`) (`True_pi`).
  tc-Decision (app [R, X, Y]) 
    (app [`decide_rel`, A, B, R, P, X, Y]) :-
    tc-RelDecision A B R P.
  tc-RelDecision (app [`sig`, A, P]) 
    (app [`sig`, A, P]) 
    (app [`eq`, app [`sig`, A, P]]) 
    (app [`sig_eq_dec`, A, P, P1, P2]) :-
    pi-decl c0 `x` A =>
      tc-ProofIrrel (app [P, c0]) (app [P1, c0]), 
    tc-RelDecision A A (app [`eq`, A]) P2.
\end{minted}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}
